---
title: "Confounding and Interaction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Confounding and Interaction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Confounding and Interaction

Confounding asks whether adjustment changes the exposure effect. Interaction
asks whether the exposure effect differs across another variable.

These checks are screening aids for viewing and organising results. Use DAGs,
subject-matter knowledge, and study design to decide which variables are
confounders or effect modifiers. Automated change-in-estimate and interaction
checks should not be used as the sole basis for model adjustment.

## Which Function Should I Use?

| Question | Use | Why |
|---|---|---|
| Could these candidate variables be confounders or effect modifiers? | `identify_confounder()` | Screens crude, adjusted, Mantel-Haenszel, and interaction signals together. |
| Does this planned interaction term improve the model? | `interaction_models()` | Compares models with and without `exposure:effect_modifier` using LRT or Wald tests. |
| Do I need a Mantel-Haenszel estimate? | `identify_confounder(method = mh)` or `identify_confounder(method = both)` | MH is a stratified pooled estimate for eligible binary/categorical settings, not a formal interaction-term test. |

A practical workflow is:

1. Use DAGs, prior literature, and study design to list important variables.
2. Use `identify_confounder()` to organise screening evidence for candidate
   confounders or effect modifiers.
3. Use `interaction_models()` when you have a planned interaction hypothesis.
4. If interaction is important, consider stratified reporting with
   `stratified_uni_reg()` or `stratified_multi_reg()`.

```{r ci-setup, message=FALSE, warning=FALSE}
library(gtregression)
library(dplyr)

data("data_birthwt", package = "gtregression")

birthwt_data <- data_birthwt |>
  mutate(
    race = factor(race, levels = c(1, 2, 3),
                  labels = c("White", "Black", "Other")),
    smoke = factor(smoke, levels = c(0, 1), labels = c("No", "Yes")),
    ht = factor(ht, levels = c(0, 1), labels = c("No", "Yes")),
    low = factor(low, levels = c(0, 1), labels = c("Normal BW", "Low BW"))
  )
```

## Identify Confounders

Use `method = change` for the model-based change-in-estimate method. Use
`method = mh` or `method = both` when Mantel-Haenszel is appropriate.
The output is intentionally tidy rather than publication-first.

```{r confounder, message=FALSE, warning=FALSE}
confounder_check <- identify_confounder(
  data = birthwt_data,
  outcome = "low",
  exposure = "smoke",
  potential_confounder = c("race", "ht"),
  approach = logit,
  method = both
)

confounder_check$summary
```

## Test Interaction

`interaction_models()` compares models with and without the interaction term.
It is deliberately model-based and uses `LRT` or `Wald`, not
Mantel-Haenszel. Use it when the interaction term is planned or supported by
clinical, causal, or subject-matter reasoning.

```{r interaction, message=FALSE, warning=FALSE}
interaction_check <- interaction_models(
  data = birthwt_data,
  outcome = "low",
  exposure = "smoke",
  effect_modifier = "race",
  covariates = c("age", "lwt"),
  approach = logit,
  test = LRT,
  format = gt
)

interaction_check$table
```

## What To Inspect

- `identify_confounder()`: `$summary`, `$table`, and `$details`.
- `interaction_models()`: `$summary`, `$table`, `$p_value`, `$decision`, and
  fitted model objects.
- Use subject-matter knowledge with these outputs. The functions support
  interpretation; they do not replace the study design.
