---
title: "Customize, Merge, and Export"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customize, Merge, and Export}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Customize, Merge, and Export

Make the output look like it belongs in the final report. Rename labels, merge
tables, and save tables or plots.

```{r export-setup, message=FALSE, warning=FALSE}
library(gtregression)
library(dplyr)

data("data_birthwt", package = "gtregression")

birthwt_data <- data_birthwt |>
  mutate(
    race = factor(race, levels = c(1, 2, 3),
                  labels = c("White", "Black", "Other")),
    smoke = factor(smoke, levels = c(0, 1), labels = c("No", "Yes")),
    ht = factor(ht, levels = c(0, 1), labels = c("No", "Yes")),
    ui = factor(ui, levels = c(0, 1), labels = c("No", "Yes")),
    low = factor(low, levels = c(0, 1), labels = c("Normal BW", "Low BW"))
  )

birthwt_desc <- descriptive_table(
  birthwt_data,
  exposures = c("age", "lwt", "race", "smoke", "ht", "ui"),
  by = "low"
)
birthwt_uni <- uni_reg(
  birthwt_data,
  outcome = "low",
  exposures = c("age", "lwt", "race", "smoke", "ht", "ui"),
  approach = "logit"
)
birthwt_multi <- multi_reg(
  birthwt_data,
  outcome = "low",
  exposures = c("smoke", "ht", "ui"),
  adjust_for = c("age", "lwt", "race"),
  approach = "logit"
)
```

## Customize Labels

```{r modify-table, message=FALSE, warning=FALSE}
birthwt_custom <- modify_table(
  birthwt_multi,
  variable_labels = c(
    smoke = "Smoking during pregnancy",
    ht = "Hypertension",
    ui = "Uterine irritability"
  ),
  header_labels = c(estimate = "Adjusted OR", p.value = "P"),
  caption = "Adjusted regression for low birth weight"
)

birthwt_custom$table
```

## Merge Tables

```{r merge-tables, message=FALSE, warning=FALSE}
birthwt_merged <- merge_tables(
  birthwt_desc,
  birthwt_uni,
  birthwt_multi,
  spanners = c("Descriptive", "Crude", "Adjusted")
)

birthwt_merged$table
```

## Save Outputs

When no destination is supplied, save helpers use a temporary folder.

```{r save-examples, eval=FALSE}
save_table(birthwt_merged, filename = "birthwt-table", format = "html")

birthwt_plot <- plot_reg(
  birthwt_multi,
  title = "Adjusted Regression for Low Birth Weight"
)

save_plot(birthwt_plot, filename = "birthwt-forest", format = "png")
```

## Word Reports

Use `format = "flextable"` when the destination is Word.

```{r word-example, eval=FALSE}
birthwt_multi_ft <- multi_reg(
  birthwt_data,
  outcome = "low",
  exposures = c("smoke", "ht", "ui"),
  adjust_for = c("age", "lwt", "race"),
  approach = "logit",
  format = "flextable"
)

save_docx(
  tables = list(birthwt_multi_ft),
  filename = "birthwt-report",
  titles = "Adjusted Regression"
)
```

## What To Inspect

- `modify_table()`: changed labels, caption, and caveat.
- `merge_tables()`: `$table`, `$table_display`, and `$footnotes`.
- `save_table()`, `save_plot()`, `save_docx()`: returned file paths.
