---
title: "Customize, Merge, and Export"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customize, Merge, and Export}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Customize, Merge, and Export

Make the output look like it belongs in the final report. Rename labels, merge
tables, and save tables or plots.

```{r export-setup, message=FALSE, warning=FALSE}
library(gtregression)
library(dplyr)

data("data_birthwt", package = "gtregression")

birthwt_data <- data_birthwt |>
  mutate(
    race = factor(race, levels = c(1, 2, 3),
                  labels = c("White", "Black", "Other")),
    smoke = factor(smoke, levels = c(0, 1), labels = c("No", "Yes")),
    ht = factor(ht, levels = c(0, 1), labels = c("No", "Yes")),
    ui = factor(ui, levels = c(0, 1), labels = c("No", "Yes")),
    low = factor(low, levels = c(0, 1), labels = c("Normal BW", "Low BW"))
  )

birthwt_exposures <- c("age", "lwt", "race", "smoke", "ht", "ui")

attr(birthwt_data$age, "label") <- "Maternal age"
attr(birthwt_data$lwt, "label") <- "Maternal weight"
attr(birthwt_data$race, "label") <- "Maternal race"
attr(birthwt_data$smoke, "label") <- "Smoking during pregnancy"
attr(birthwt_data$ht, "label") <- "Hypertension"
attr(birthwt_data$ui, "label") <- "Uterine irritability"

birthwt_desc <- descriptive_table(
  birthwt_data,
  exposures = birthwt_exposures,
  by = low
)
birthwt_uni <- uni_reg(
  birthwt_data,
  outcome = low,
  exposures = birthwt_exposures,
  approach = logit
)
birthwt_multi <- multi_reg(
  birthwt_data,
  outcome = low,
  exposures = c("smoke", "ht", "ui"),
  adjust_for = c("age", "lwt", "race"),
  approach = logit
)
```

## Customize Labels

If labels already live on the data, `gtregression` uses them automatically.
`modify_table()` is still useful for journal-specific wording, compact headers,
captions, and caveats. Use raw variable names on the left side of
`variable_labels` and `level_labels`; this keeps customisation stable even when
the visible table already shows prettier labels.

Footnotes and caveats are styled compactly by default across flextable and gt
outputs, which keeps abbreviation notes and adjustment notes readable without
making final tables unnecessarily tall.

```{r modify-table, message=FALSE, warning=FALSE}
birthwt_custom <- modify_table(
  birthwt_multi,
  variable_labels = c(
    smoke = "Smoked during pregnancy",
    ht = "History of hypertension",
    ui = "Uterine irritability"
  ),
  level_labels = list(
    smoke = c(Yes = "Smoker"),
    ht = c(Yes = "Hypertensive")
  ),
  header_labels = c(estimate = "Adjusted OR", p.value = "P"),
  caption = "Adjusted regression for low birth weight",
  caveat = "Adjusted for maternal age, maternal weight, and maternal race."
)

birthwt_custom$table
```

## Merge Tables

`merge_tables()` combines descriptive, crude, and adjusted results. Matching is
based on the original variable names, so merged tables remain aligned even when
the visible labels differ across input tables.

```{r merge-tables, message=FALSE, warning=FALSE}
birthwt_merged <- merge_tables(
  birthwt_desc,
  birthwt_uni,
  birthwt_multi,
  spanners = c("Descriptive", "Crude", "Adjusted")
)

birthwt_merged$table
```

The merged table can be polished after merging too.

```{r merged-polish, message=FALSE, warning=FALSE}
birthwt_merged_paper <- modify_table(
  birthwt_merged,
  variable_labels = c(
    age = "Maternal age",
    lwt = "Maternal weight",
    race = "Maternal race",
    smoke = "Smoking during pregnancy",
    ht = "Hypertension",
    ui = "Uterine irritability"
  ),
  caption = "Clinical profile and regression estimates for low birth weight",
  caveat = "Adjusted estimates are adjusted for maternal age, maternal weight, and maternal race."
)

birthwt_merged_paper$table
```

## Save Outputs

When no directory is supplied, save helpers use `tempdir()`. This keeps examples
and tests CRAN-safe while still returning the file path invisibly.

```{r save-examples, eval=FALSE}
table_path <- save_table(
  birthwt_merged_paper,
  filename = "birthwt-table",
  format = html
)

birthwt_plot <- plot_reg(
  birthwt_multi,
  title = "Adjusted Regression for Low Birth Weight"
)

plot_path <- save_plot(
  birthwt_plot,
  filename = "birthwt-forest",
  format = png
)
```

## Word Reports

`flextable` is the default table engine, so Word export works naturally. If a
table was created as `format = gt`, save it as HTML/PDF or recreate it with
`format = flextable` before sending it to `save_docx()`. Wide tables are fitted
to a standard Word page by default; use `table_width` when your document has
different margins or landscape orientation.

```{r word-example, eval=FALSE}
birthwt_multi_ft <- multi_reg(
  birthwt_data,
  outcome = low,
  exposures = c("smoke", "ht", "ui"),
  adjust_for = c("age", "lwt", "race"),
  approach = logit,
  format = flextable
)

docx_path <- save_docx(
  tables = list(birthwt_multi_ft),
  filename = "birthwt-report",
  titles = "Adjusted Regression",
  table_width = 6.5
)
```

## What To Inspect

- `modify_table()`: changed labels, caption, and caveat.
- `merge_tables()`: `$table`, `$table_display`, and `$footnotes`.
- `save_table()`, `save_plot()`, `save_docx()`: invisibly returned file paths.
- Raw variable names remain available for relabelling, merging, and testing even
  when display labels are used.
