---
title: "Descriptive Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Descriptive Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Descriptive Tables

Start with a table that people can actually read. `descriptive_table()` creates
publication-ready summaries for continuous, categorical, and mixed exposure
sets.

```{r desc-setup, message=FALSE, warning=FALSE}
library(gtregression)
library(dplyr)

data("data_birthwt", package = "gtregression")

birthwt_data <- data_birthwt |>
  mutate(
    race = factor(race, levels = c(1, 2, 3),
                  labels = c("White", "Black", "Other")),
    smoke = factor(smoke, levels = c(0, 1), labels = c("No", "Yes")),
    ht = factor(ht, levels = c(0, 1), labels = c("No", "Yes")),
    ui = factor(ui, levels = c(0, 1), labels = c("No", "Yes")),
    low = factor(low, levels = c(0, 1), labels = c("Normal BW", "Low BW")),
    ptl_cat = factor(ifelse(ptl > 0, "Yes", "No"), levels = c("No", "Yes")),
    ftv_cat = factor(case_when(
      ftv == 0 ~ "None",
      ftv == 1 ~ "One",
      ftv >= 2 ~ "Two or more"
    ), levels = c("None", "One", "Two or more"))
  )

birthwt_exposures <- c(
  "age", "lwt", "race", "smoke", "ht", "ui", "ptl_cat", "ftv_cat"
)
```

## Column Percentages

Use column percentages when the table is grouped by outcome or another column.

```{r desc-column, message=FALSE, warning=FALSE}
desc_column <- descriptive_table(
  data = birthwt_data,
  exposures = birthwt_exposures,
  by = "low",
  percent = column,
  show_overall = last,
  theme = clinical
)

desc_column$table
```

## Row Percentages

Use row percentages when the question is how each exposure level is distributed
across groups.

```{r desc-row, message=FALSE, warning=FALSE}
descriptive_table(
  data = birthwt_data,
  exposures = birthwt_exposures,
  by = "low",
  percent = row,
  show_overall = first,
  show_missing = no,
  theme = striped
)$table
```

## Word-Friendly Output

Switch to `format = flextable` when the next stop is a Word report.

```{r desc-flex, message=FALSE, warning=FALSE}
descriptive_table(
  data = birthwt_data,
  exposures = birthwt_exposures,
  by = "low",
  percent = column,
  format = flextable
)$table
```

## What To Inspect

- `$table`: rendered `gt` or `flextable` output.
- `$table_body`: clean data behind the table.
- `$format`: output format used by the table builder.
