---
title: "Diagnostics and Model Selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostics and Model Selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Diagnostics and Model Selection

Before the final table goes into a manuscript, check the model. These helpers
keep diagnostics close to the regression workflow.

```{r diag-setup, message=FALSE, warning=FALSE}
library(gtregression)
library(dplyr)

data("data_birthwt", package = "gtregression")

birthwt_data <- data_birthwt |>
  mutate(
    race = factor(race, levels = c(1, 2, 3),
                  labels = c("White", "Black", "Other")),
    smoke = factor(smoke, levels = c(0, 1), labels = c("No", "Yes")),
    ht = factor(ht, levels = c(0, 1), labels = c("No", "Yes")),
    ui = factor(ui, levels = c(0, 1), labels = c("No", "Yes")),
    low = factor(low, levels = c(0, 1), labels = c("Normal BW", "Low BW")),
    ptl_cat = factor(ifelse(ptl > 0, "Yes", "No"), levels = c("No", "Yes"))
  )

exposures <- c("age", "lwt", "race", "smoke", "ht", "ui", "ptl_cat")
```

## Convergence

```{r diag-convergence, message=FALSE, warning=FALSE}
check_convergence(
  data = birthwt_data,
  exposures = exposures,
  outcome = "low",
  approach = logit,
  multivariate = TRUE,
  format = gt
)
```

## Collinearity

```{r diag-collinearity, message=FALSE, warning=FALSE}
birthwt_multi <- multi_reg(
  data = birthwt_data,
  outcome = "low",
  exposures = exposures,
  approach = logit
)

check_collinearity(birthwt_multi, format = gt)
```

## Stepwise Model Selection

```{r diag-select, message=FALSE, warning=FALSE}
selected <- select_models(
  data = birthwt_data,
  outcome = "low",
  exposures = exposures,
  approach = logit,
  direction = forward,
  format = gt
)

selected$table
```

## What To Inspect

- `check_convergence()`: convergence status and maximum fitted probabilities.
  Use `format = gt` or `format = flextable` for viewing tables.
- `check_collinearity()`: VIF and interpretation. Nested model outputs keep
  their list structure when formatted.
- `select_models()`: `$results_table`, `$best_model`, `$all_models`, and
  `$table` when `format = gt` or `format = flextable`.
