---
title: "Diagnostics and Model Selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostics and Model Selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Diagnostics and Model Selection

Before the final table goes into a manuscript, check the model. These helpers
keep diagnostics close to the regression workflow. They are screening aids:
interpret them with the study design, clinical or subject-matter judgement, and
the diagnostics from the fitted model.

```{r diag-setup, message=FALSE, warning=FALSE}
library(gtregression)
library(dplyr)

data("data_birthwt", package = "gtregression")

birthwt_data <- data_birthwt |>
  mutate(
    race = factor(race, levels = c(1, 2, 3),
                  labels = c("White", "Black", "Other")),
    smoke = factor(smoke, levels = c(0, 1), labels = c("No", "Yes")),
    ht = factor(ht, levels = c(0, 1), labels = c("No", "Yes")),
    ui = factor(ui, levels = c(0, 1), labels = c("No", "Yes")),
    low = factor(low, levels = c(0, 1), labels = c("Normal BW", "Low BW")),
    ptl_cat = factor(ifelse(ptl > 0, "Yes", "No"), levels = c("No", "Yes"))
  )

exposures <- c("age", "lwt", "race", "smoke", "ht", "ui", "ptl_cat")
```

## Convergence Screening

Use `check_convergence()` before interpreting model estimates, especially for
log-binomial and small or sparse binary-outcome models. A non-converged model is
a fitting warning, not a finding.

```{r diag-convergence, message=FALSE, warning=FALSE}
check_convergence(
  data = birthwt_data,
  exposures = exposures,
  outcome = low,
  approach = logit,
  multivariate = TRUE,
  format = gt
)
```

For risk-ratio workflows, this same check helps users decide whether a
log-binomial model fitted cleanly or whether a robust Poisson approach may be a
more practical sensitivity analysis.

```{r diag-convergence-logbin, message=FALSE, warning=FALSE}
check_convergence(
  data = birthwt_data,
  exposures = c("smoke", "ht", "ui", "ptl_cat"),
  outcome = low,
  approach = logbinomial,
  multivariate = TRUE,
  format = flextable
)
```

## Collinearity Screening

`check_collinearity()` reports VIF-style diagnostics for multivariable models.
High VIF values are prompts to inspect coding, overlap between predictors, and
the scientific purpose of the model.

```{r diag-collinearity, message=FALSE, warning=FALSE}
birthwt_multi <- multi_reg(
  data = birthwt_data,
  outcome = low,
  exposures = exposures,
  approach = logit
)

check_collinearity(birthwt_multi, format = gt)
```

Adjusted-mode `multi_reg()` objects contain one model per exposure. The
collinearity output keeps that list structure so each model can be inspected
separately.

```{r diag-collinearity-adjusted, message=FALSE, warning=FALSE}
birthwt_adjusted <- multi_reg(
  data = birthwt_data,
  outcome = low,
  exposures = c("smoke", "ht", "ui", "ptl_cat"),
  adjust_for = c("age", "lwt", "race"),
  approach = logit
)

check_collinearity(birthwt_adjusted, format = tibble)
```

## Model Fit Plots

`plot_model_fit()` turns fitted models into quick diagnostic plots. It accepts
raw `lm()` and `glm()` objects, and it also works with models saved inside
`uni_reg()` and `multi_reg()` results.

For logistic regression, the calibration plot compares predicted probabilities
with observed event proportions. Points close to the diagonal line suggest that
the model predictions are reasonably aligned with the observed data. Calibration
is usually most useful for multivariable models because predicted probabilities
vary across many people.

```{r diag-fit-logistic, message=FALSE, warning=FALSE}
plot_model_fit(
  birthwt_multi,
  type = calibration,
  bins = 6
)
```

When a `uni_reg()` object contains several models, use `model_name` to choose
the exposure you want to inspect. For a simple binary exposure, calibration may
only show two points because the model has only two fitted probabilities; in
that situation, residual and influence plots are usually more useful.

```{r diag-fit-uni, message=FALSE, warning=FALSE}
birthwt_uni <- uni_reg(
  data = birthwt_data,
  outcome = low,
  exposures = c("age", "lwt", "smoke"),
  approach = logit
)

plot_model_fit(
  birthwt_uni,
  model_name = smoke,
  type = residual
)
```

For logistic models, residual plots often form two visible bands. That is a
normal consequence of a 0/1 outcome and should be interpreted as a screening
plot rather than a linear-model residual plot.

For linear regression, `type = all` shows the classic residual, Q-Q,
scale-location, and Cook's distance views.

```{r diag-fit-linear, message=FALSE, warning=FALSE}
fit_lm <- lm(bwt ~ age + lwt, data = birthwt_data)
plot_model_fit(fit_lm)
```

## Proportional Hazards Screening

For Cox models, use `check_ph()` before treating hazard ratios as final. It
reports Schoenfeld residual tests from `survival::cox.zph()`, including a global
test. Small p-values suggest possible non-proportional hazards and should be
reviewed with plots, follow-up pattern, and clinical judgement.

```{r diag-ph-setup, message=FALSE, warning=FALSE}
data("data_lungcancer", package = "gtregression")

lung_data <- data_lungcancer |>
  mutate(
    trt = factor(trt, levels = c(1, 2),
                 labels = c("Standard treatment", "Test treatment")),
    prior = factor(prior, levels = c(0, 10), labels = c("No", "Yes"))
  )

cox_fit <- cox_reg(
  data = lung_data,
  time = time,
  event = status,
  exposures = c(trt, celltype, prior),
  adjust_for = c(age, karno)
)
```

```{r diag-ph-table, message=FALSE, warning=FALSE}
check_ph(cox_fit, format = gt)
```

Use `format = tibble` when you want to inspect or filter the diagnostic results.

```{r diag-ph-tibble, message=FALSE, warning=FALSE}
check_ph(cox_fit, transform = rank, format = tibble)
```

## Stepwise Model Selection

`select_models()` compares candidate models step by step. It is useful for
exploration, teaching, and sensitivity checks. It should not replace a planned
model based on study design or a causal framework.

```{r diag-select, message=FALSE, warning=FALSE}
selected <- select_models(
  data = birthwt_data,
  outcome = low,
  exposures = exposures,
  approach = logit,
  direction = forward,
  format = gt
)

selected$table
```

The selected direction is recorded in the formatted table footer. Backward and
both-direction searches are available using the same interface.

```{r diag-select-other-directions, message=FALSE, warning=FALSE}
select_models(
  data = birthwt_data,
  outcome = low,
  exposures = exposures,
  approach = logit,
  direction = backward,
  format = tibble
)$results_table

select_models(
  data = birthwt_data,
  outcome = low,
  exposures = exposures,
  approach = logit,
  direction = both,
  format = tibble
)$results_table
```

## What To Inspect

- `check_convergence()`: convergence status and maximum fitted probabilities.
  Use `format = gt` or `format = flextable` for viewing tables.
- `check_collinearity()`: VIF and interpretation. Nested model outputs keep
  their list structure when formatted.
- `plot_model_fit()`: residual, calibration, observed-versus-predicted, and
  influence plots for `lm`/`glm` models and stored `uni_reg()` / `multi_reg()`
  fitted models.
- `check_ph()`: Schoenfeld residual proportional hazards tests for Cox models,
  including term-level and global tests.
- `select_models()`: `$results_table`, `$best_model`, `$all_models`, and
  `$direction`; `$table` is added when `format = gt` or `format = flextable`.
