---
title: "Using the gtregression GUI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GUI Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

The gtregression GUI is a point-and-click Shiny app that wraps the core functions of the package—building descriptive tables, fitting regression models, checking diagnostics, and exporting publication-ready outputs. It’s designed to feel familiar to SPSS users while keeping everything reproducible (the app can show and download the R code it runs).

This vignette gives you:

-   how to launch the GUI,

-   a quick tour of each panel,

-   an end-to-end example using the included Pima data, and

-   programmatic equivalents if you prefer code.

## Installation & Launch

```{r}
# Install from CRAN (if available)
# install.packages("gtregression")

# Or install the development version from GitHub
# install.packages("devtools")
# devtools::install_github("ThinkDenominator/gtregression")

library(gtregression)

# Launch the GUI (opens in your browser)
# (kept non-evaluated for vignette builds)
```

```text
# NOT RUN {
gtregression_app() 
# }
```

## Suggested packages

The GUI will use (and gently prompt for) a few optional packages to enable specific features:

-   shiny, shinythemes, shinyjs, DT

-   gt, gtsummary, ggplot2, patchwork/cowplot

-   readxl and/or rio for imports

-   officer/flextable for DOCX exports

-   MASS, sandwich, lmtest for certain model types

You don’t need to install all of them to get started—the app disables features that aren’t available.

## A quick tour of the GUI 
### Import Data

Upload a CSV or XLSX file, or pick a built-in dataset from the dropdown.

The Data Preview shows the first 10 rows (you can switch to all rows via the “Preview all rows” toggle in the Import module).

### Descriptive Tables

Choose one or more exposures to summarize.

Optionally stratify by another variable (e.g., an outcome).

Choose percentage type (column/row) and whether to add an overall column.

Click Run Descriptive Table, view the formatted gt table, and export to DOCX/HTML/PDF/PNG.

A Reproducible Code tab shows exactly how to recreate the table in plain R.

### Regression Analysis

Tabs inside:

Univariate: runs a separate model for each exposure.

Multivariable: fits a single model with all selected exposures.

Stratified: runs the selected approach within levels of a stratifier (univariate or multivariable).

Combine Results: merge tables (with spanners “Descriptive”, “Univariate”, “Multivariable”) and optionally combine plots.

Each sub-tab includes Table, Plot, Model Summary, and Reproducible Code (downloadable as an .R script).

### Advanced

Select Models: stepwise model selection (forward/backward/both) across allowed approaches.

Identify Confounders: compare crude vs. adjusted effect for a single exposure + potential confounder at a user-set threshold (default 10%).

Interaction Models: compare models with/without the interaction term; displays the same console output as the underlying function.

Convergence: fit and report convergence diagnostics (where applicable).
