---
title: "Regression Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Go from fitted models to publication-ready tables without hand-formatting
effect estimates. `gtregression` supports logistic, log-binomial, Poisson,
robust Poisson, negative binomial, Cox survival, parametric survival, and
linear regression.

```{r reg-setup, message=FALSE, warning=FALSE}
library(gtregression)
library(dplyr)

data("data_birthwt", package = "gtregression")

birthwt_data <- data_birthwt |>
  mutate(
    race = factor(race, levels = c(1, 2, 3),
                  labels = c("White", "Black", "Other")),
    smoke = factor(smoke, levels = c(0, 1), labels = c("No", "Yes")),
    ht = factor(ht, levels = c(0, 1), labels = c("No", "Yes")),
    ui = factor(ui, levels = c(0, 1), labels = c("No", "Yes")),
    low = factor(low, levels = c(0, 1), labels = c("Normal BW", "Low BW")),
    ptl_cat = factor(ifelse(ptl > 0, "Yes", "No"), levels = c("No", "Yes")),
    ftv_cat = factor(case_when(
      ftv == 0 ~ "None",
      ftv == 1 ~ "One",
      ftv >= 2 ~ "Two or more"
    ), levels = c("None", "One", "Two or more"))
  )

birthwt_exposures <- c(
  "age", "lwt", "race", "smoke", "ht", "ui", "ptl_cat", "ftv_cat"
)

attr(birthwt_data$age, "label") <- "Maternal age"
attr(birthwt_data$lwt, "label") <- "Maternal weight"
attr(birthwt_data$race, "label") <- "Maternal race"
attr(birthwt_data$smoke, "label") <- "Smoking during pregnancy"
attr(birthwt_data$ht, "label") <- "Hypertension"
attr(birthwt_data$ui, "label") <- "Uterine irritability"
attr(birthwt_data$ptl_cat, "label") <- "Previous preterm labour"
attr(birthwt_data$ftv_cat, "label") <- "First trimester visits"
```

## Univariable Models

`uni_reg()` fits one model per exposure and returns a table ready for reports.
Variable labels set with `attr(x, "label")` or `labelled::var_label()` are used
automatically in the displayed table, while raw column names remain in
`$table_body`.

```{r uni-logit, message=FALSE, warning=FALSE}
birthwt_uni <- uni_reg(
  data = birthwt_data,
  outcome = "low",
  exposures = birthwt_exposures,
  approach = "logit",
  theme = clinical
)

birthwt_uni$table
```

## Multivariable Models

With `adjust_for = NULL`, `multi_reg()` fits all supplied exposures in one
multivariable model. This is the usual fully adjusted model when every exposure
listed should appear in the same formula.

```{r multi-full, message=FALSE, warning=FALSE}
birthwt_full <- multi_reg(
  data = birthwt_data,
  outcome = low,
  exposures = c("age", "lwt", "race", "smoke", "ht", "ui"),
  approach = logit,
  theme = clinical
)

birthwt_full$table
```

## Exposure-Specific Adjusted Models

Use `adjust_for` when you want one adjusted model per exposure, each adjusted for
the same core covariate set. This is useful for screening several clinically
important exposures while keeping the adjustment strategy explicit.

```{r multi-adjusted, message=FALSE, warning=FALSE}
birthwt_multi <- multi_reg(
  data = birthwt_data,
  outcome = low,
  exposures = c("smoke", "ht", "ui", "ptl_cat", "ftv_cat"),
  adjust_for = c("age", "lwt", "race"),
  approach = logit,
  theme = striped
)

birthwt_multi$table
```

The adjustment variables are recorded in a compact table footnote, so the result
is ready for manuscript-style reporting without making the table unnecessarily
tall.

## Optional Model Statistics

Publication tables should stay readable. When you need model-fit information,
set `model_stats = TRUE` and inspect the returned object's `$model_stats`
element. This keeps AIC, BIC, log-likelihood, deviance, pseudo R-squared, and
linear-model R-squared values available without adding clutter to the main
table.

```{r model-stats, message=FALSE, warning=FALSE}
birthwt_uni_stats <- uni_reg(
  data = birthwt_data,
  outcome = low,
  exposures = birthwt_exposures,
  approach = logit,
  model_stats = TRUE
)

birthwt_uni_stats$model_stats
```

For adjusted mode, `multi_reg()` returns one row of statistics per adjusted
exposure-specific model.

```{r adjusted-model-stats, message=FALSE, warning=FALSE}
birthwt_multi_stats <- multi_reg(
  data = birthwt_data,
  outcome = low,
  exposures = c("smoke", "ht", "ui", "ptl_cat"),
  adjust_for = c("age", "lwt", "race"),
  approach = logit,
  model_stats = TRUE
)

birthwt_multi_stats$model_stats
```

## Other Effect Measures

Switch the `approach` to change the estimand.

```{r reg-risk-ratio, message=FALSE, warning=FALSE}
uni_reg(
  data = birthwt_data,
  outcome = low,
  exposures = c("smoke", "ht", "ui", "ptl_cat"),
  approach = logbinomial
)$table
```

Use `approach = firth` when a binary-outcome logistic model has sparse cells,
very wide intervals, or separation concerns. The output remains an odds-ratio
table, but the model is fitted with Firth penalized logistic regression.

The built-in `data_endometrial` dataset is a useful teaching example because
neovascularization is completely absent among low-grade cases, a pattern that
can make ordinary logistic regression unstable.

```{r reg-firth, message=FALSE, warning=FALSE}
data("data_endometrial", package = "gtregression")

endometrial_data <- data_endometrial |>
  mutate(
    HG = factor(HG, levels = c(0, 1),
                labels = c("Low grade", "High grade")),
    NV = factor(NV, levels = c(0, 1), labels = c("Absent", "Present"))
  )

multi_reg(
  data = endometrial_data,
  outcome = HG,
  exposures = c(NV, PI, EH),
  approach = firth
)$table
```

## Cox Survival Models

`cox_reg()` uses direct `time` and `event` arguments and returns
hazard ratios. Without `adjust_for`, the table shows crude HRs. With
`adjust_for`, the table shows adjusted HRs.

```{r cox-setup, message=FALSE, warning=FALSE}
data("data_lungcancer", package = "gtregression")

lung_data <- data_lungcancer |>
  mutate(
    trt = factor(trt, levels = c(1, 2),
                 labels = c("Standard treatment", "Test treatment")),
    prior = factor(prior, levels = c(0, 10), labels = c("No", "Yes"))
  )

attr(lung_data$trt, "label") <- "Treatment group"
attr(lung_data$celltype, "label") <- "Cancer cell type"
attr(lung_data$karno, "label") <- "Karnofsky performance score"
attr(lung_data$age, "label") <- "Age"
attr(lung_data$prior, "label") <- "Prior therapy"
```

```{r cox-crude, message=FALSE, warning=FALSE}
lung_hr <- cox_reg(
  data = lung_data,
  time = time,
  event = status,
  exposures = c("trt", "celltype", "karno", "age"),
  theme = clinical
)

lung_hr$table
```

```{r cox-adjusted, message=FALSE, warning=FALSE}
lung_adj_hr <- cox_reg(
  data = lung_data,
  time = time,
  event = status,
  exposures = c(trt, celltype, prior),
  adjust_for = c(age, karno),
  model_stats = TRUE,
  theme = striped
)

lung_adj_hr$table
lung_adj_hr$model_stats
```

Use `interaction = exposure*modifier` for planned effect modification. In the
default exposure-by-exposure workflow, provide one exposure.

```{r cox-interaction, message=FALSE, warning=FALSE}
cox_interaction <- cox_reg(
  data = lung_data,
  time = time,
  event = status,
  exposures = trt,
  adjust_for = c(age, karno),
  interaction = trt*prior
)

cox_interaction$table
```

## Parametric Survival Models

`surv_reg()` uses the same `time`, `event`, `exposures`, and `adjust_for`
grammar as `cox_reg()`, but fits parametric survival models with
`survival::survreg()`. The table reports time ratios rather than hazard ratios.
A time ratio above 1 suggests longer survival time; below 1 suggests shorter
survival time, conditional on the chosen distribution.

```{r surv-crude, message=FALSE, warning=FALSE}
lung_time_ratio <- surv_reg(
  data = lung_data,
  time = time,
  event = status,
  exposures = c("trt", "celltype", "karno", "age"),
  distribution = weibull,
  theme = clinical
)

lung_time_ratio$table
```

```{r surv-adjusted, message=FALSE, warning=FALSE}
lung_adj_time_ratio <- surv_reg(
  data = lung_data,
  time = time,
  event = status,
  exposures = c(trt, celltype, prior),
  adjust_for = c(age, karno),
  distribution = lognormal,
  model_stats = TRUE,
  theme = striped
)

lung_adj_time_ratio$table
lung_adj_time_ratio$model_stats
```

```{r surv-interaction, message=FALSE, warning=FALSE}
surv_interaction <- surv_reg(
  data = lung_data,
  time = time,
  event = status,
  exposures = trt,
  adjust_for = c(age, karno),
  interaction = trt*prior,
  distribution = weibull
)

surv_interaction$table
```

## Continuous Outcomes

Linear regression outputs beta coefficients and keeps diagnostics under
`$reg_check`.

```{r reg-linear, message=FALSE, warning=FALSE}
birthwt_linear <- multi_reg(
  data = birthwt_data,
  outcome = bwt,
  exposures = c("age", "lwt", "race", "smoke", "ht", "ui"),
  approach = linear
)

birthwt_linear$table
```

## What To Inspect

- `$table`: publication-ready table.
- `$table_body`: numeric estimates behind the display.
- `$models`: fitted model objects.
- `$model_summaries`: model-level summaries.
- `$model_stats`: optional model-fit statistics when `model_stats = TRUE`.
- `$variable_labels`: labels used in publication output.
- `$reg_check`: diagnostics for linear models.
