---
title: "Stratified Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stratified Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Stratified regression repeats the analysis inside each subgroup and places the
results side by side. It is useful when the same association may look different
across groups.

```{r strata-setup, message=FALSE, warning=FALSE}
library(gtregression)
library(dplyr)

data("data_birthwt", package = "gtregression")

birthwt_data <- data_birthwt |>
  mutate(
    race = factor(race, levels = c(1, 2, 3),
                  labels = c("White", "Black", "Other")),
    smoke = factor(smoke, levels = c(0, 1), labels = c("No", "Yes")),
    ht = factor(ht, levels = c(0, 1), labels = c("No", "Yes")),
    ui = factor(ui, levels = c(0, 1), labels = c("No", "Yes")),
    low = factor(low, levels = c(0, 1), labels = c("Normal BW", "Low BW")),
    ptl_cat = factor(ifelse(ptl > 0, "Yes", "No"), levels = c("No", "Yes"))
  )

attr(birthwt_data$age, "label") <- "Maternal age"
attr(birthwt_data$lwt, "label") <- "Maternal weight"
attr(birthwt_data$smoke, "label") <- "Smoking during pregnancy"
attr(birthwt_data$ht, "label") <- "Hypertension"
attr(birthwt_data$ui, "label") <- "Uterine irritability"
attr(birthwt_data$ptl_cat, "label") <- "Previous preterm labour"
```

## Describe by Stratum

Start with a descriptive table by the stratifying variable. This is the companion
table for the stratified regression: it helps users see the size and clinical
profile of each subgroup before fitting stratum-specific models.

```{r strata-desc, message=FALSE, warning=FALSE}
strata_desc <- descriptive_table(
  data = birthwt_data,
  exposures = c("age", "lwt", "smoke", "ht", "ui", "ptl_cat"),
  by = race,
  percent = column,
  show_overall = last,
  theme = clinical
)

strata_desc$table
```

## Univariable by Stratum

`stratified_uni_reg()` fits one model per exposure inside each stratum. The
result is a single wide table, with one spanner per stratum.

```{r strata-uni, message=FALSE, warning=FALSE}
strata_uni <- stratified_uni_reg(
  data = birthwt_data,
  outcome = low,
  exposures = c("age", "lwt", "smoke", "ht", "ui", "ptl_cat"),
  stratifier = race,
  approach = logit,
  theme = clinical
)

strata_uni$table
```

## Full Multivariable Model by Stratum

With `adjust_for = NULL`, `stratified_multi_reg()` fits one multivariable model
inside each stratum using all supplied exposures.

```{r strata-multi-full, message=FALSE, warning=FALSE}
strata_full <- stratified_multi_reg(
  data = birthwt_data,
  outcome = low,
  exposures = c("age", "lwt", "smoke", "ht", "ui", "ptl_cat"),
  stratifier = race,
  approach = logit,
  theme = clinical
)

strata_full$table
```

## Exposure-Specific Adjusted Models by Stratum

Use `adjust_for` when each exposure should be adjusted for the same variables
within each stratum. This mirrors `multi_reg(adjust_for = ...)`, but repeats the
same workflow separately inside each stratum.

```{r strata-multi, message=FALSE, warning=FALSE}
strata_multi <- stratified_multi_reg(
  data = birthwt_data,
  outcome = low,
  exposures = c("smoke", "ht", "ui", "ptl_cat"),
  stratifier = race,
  adjust_for = c("age", "lwt"),
  approach = logit,
  theme = striped
)

strata_multi$table
```

If a stratum cannot fit a model, the function skips that stratum with a warning
and continues. This is intentional: sparse strata are common in real data, and
one small subgroup should not erase the whole analysis.

## What To Inspect

- `$table`: rendered side-by-side table.
- `$table_display`: wide data used to build the table.
- `$per_stratum`: full per-stratum result objects.
- `$models`: fitted models by stratum.
- `$model_summaries`: summaries for the fitted models.
- `$variable_labels`: display labels used in the wide table.
- `$reg_check`: diagnostics for linear models.
